---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
options(tibble.print_min = 3, tibble.print_max = 3)
```

[![Travis-CI Build Status](https://travis-ci.org/echasnovski/keyholder.svg?branch=master)](https://travis-ci.org/echasnovski/keyholder)
[![Coverage Status](https://img.shields.io/codecov/c/github/echasnovski/keyholder/master.svg)](https://codecov.io/github/echasnovski/keyholder?branch=master)
[![packageversion](https://img.shields.io/badge/Package%20version-0.0.0.9000-green.svg?style=flat-square)](commits/master)

# keyholder

`keyholder` is a package for storing information (*keys*) about rows in data
frame like objects. The common use cases are to track rows of data without
modifying it and to backup and restore information about rows. This is done with
creating a class __keyed_df__ which has special attribute "keys". Keys are
updated according to changes in rows of reference data frame.

`keyholder` is designed to work tightly with
[dplyr](https://github.com/tidyverse/dplyr) package. There are methods
implemented for all one- and two-table generics from this package that update
keys properly.

## Installation

You can install `keyholder` from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("echasnovski/keyholder")
```

## Usage

```{r setup, message = FALSE}
library(dplyr)
library(keyholder)
mtcars_tbl <- mtcars %>% as_tibble()
```

### Set keys

The general agreement is that keys are always converted to [tibble](https://github.com/tidyverse/tibble). In this way one can use multiple
variables as keys by binding them.

There are two ways of creating keys:

- With assigning. The assigned object will be converted to tibble with
`as_tibble()`. To make sense it should have the same number of rows as reference
data frame. There are two functions for assigning: `keys<-` and `assign_keys`
which are basically the same. The former use more suitable for interactive use
and the latter - for piping with
[magrittr](https://github.com/tidyverse/magrittr)'s pipe operator `%>%`.

```{r set keys by assigning}
mtcars_tbl_keyed <- mtcars_tbl
keys(mtcars_tbl_keyed) <- tibble(id = 1:nrow(mtcars_tbl_keyed))

mtcars_tbl %>% assign_keys(tibble(id = 1:nrow(.)))
```

- With `key_by()`. This is similar in its design to `group_by` from
`dplyr`: it takes some columns from reference data frame and makes them keys.
It has two important options: `.add` (whether to add specified columns to
existing keys) and `.exclude` (whether to exclude specified columns from
reference data frame). Grouping is ignored.

```{r sey keys by key_by}
mtcars_tbl %>% key_by(vs, am)

mtcars_tbl %>% key_by(starts_with("c"))
```

To properly unkey object use `unkey()`.

```{r unkey}
mtcars_tbl_keyed <- mtcars_tbl %>% key_by(vs, am)

# Good
mtcars_tbl_keyed %>% unkey()

# Bad
attr(mtcars_tbl_keyed, "keys") <- NULL
mtcars_tbl_keyed
```

### Get keys

There are also two ways of extracting keys:

- With `keys()`. This function always returns a tibble. In case of no keys it
returns a tibble with number of rows as in reference data frame and zero
columns.

```{r get keys with keys}
mtcars_tbl %>% keys()

mtcars_tbl %>% key_by(vs, am) %>% keys()
```

- With `raw_keys()` which is just a wrapper for `attr(.tbl, "keys")`.

```{r get keys with raw_keys}
mtcars_tbl %>% raw_keys()

mtcars_tbl %>% key_by(vs, am) %>% raw_keys()
```

### Manipulate keys

- Remove keys with `remove_keys()`. If all keys are removed one can
automatically unkey object by setting option `.unkey` to `TRUE`.

```{r remove keys}
mtcars_tbl %>% key_by(vs, am) %>% remove_keys(vs)

mtcars_tbl %>% key_by(vs, am) %>% remove_keys(everything(), .unkey = TRUE)
```

- Restore keys with `restore_keys()`. Restoring means creating or modifying a
column in reference data frame with values taken from keys. After restoring
certain key one can remove it from keys by setting `.remove` to `TRUE`. There is
also an option `.unkey` identical to one in `remove_keys()` (which is meaningful
only in case `.remove` is `TRUE`).

```{r restore keys}
mtcars_tbl_keyed <- mtcars_tbl %>%
  key_by(vs, am) %>%
  mutate(vs = 1, am = 0)
mtcars_tbl_keyed

mtcars_tbl_keyed %>% restore_keys(vs)

mtcars_tbl_keyed %>% restore_keys(vs, .remove = TRUE)

mtcars_tbl_keyed %>% restore_keys(vs, am, .unkey = TRUE)

mtcars_tbl_keyed %>% restore_keys(vs, am, .remove = TRUE, .unkey = TRUE)
```

One important feature of `restore_keys()` is that restoring keys beats 
'not-modifying' grouping variables rule. It is made according to the ideology of
keys: they contain information about rows and by restoring you want it to be
available.

```{r restore keys grouping}
mtcars_tbl_keyed %>% group_by(vs, am)

mtcars_tbl_keyed %>% group_by(vs, am) %>% restore_keys(vs, am)
```

- Rename keys with `rename_keys()`. Renaming is done with `rename()` from
`dplyr` and so renaming format comes from it.

```{r rename keys}
mtcars_tbl %>% key_by(vs, am) %>% rename_keys(Vs = vs)
```

### Verbs from dplyr

All one- and two-table verbs from `dplyr` (with present scoped variants) support
`keyed_df`. If rows in reference data frame rearranged or removed the same
operation is doen to keys. Some functions (`summarise`, `distinct` and `do`)
unkey object.

```{r dplyr verbs}
mtcars_tbl_keyed <- mtcars_tbl %>% key_by(vs, am)

mtcars_tbl_keyed %>% select(gear, mpg)

mtcars_tbl_keyed %>% summarise(meanMPG = mean(mpg))

mtcars_tbl_keyed %>% filter(vs == 1) %>% keys()

mtcars_tbl_keyed %>% arrange_at("mpg") %>% keys()
```
